---
title: "Plastic Pollution and its Economic Drivers"
author: "Jonathan Flores Martinez, Avery Stewart, Justin Mai"
format:
  html: 
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
editor: source
embed-resources: true
execute: 
  echo: true
  warning: false
  message: false
toc: true
toc-location: right
theme:
  light: cerulean
  dark: slate
---

# The Context of our Data


## Original Data Set
This dataset was given to us... etc, the 5 whys
who where when how why
This was a data frame with information on plastic pollution. It was originally collected by the Break Free From Plastic (BFFP) group. They are an environmental activist with an interest in reducing the amount of plastic waste in the world. They focus on reducing the use of single use plastics and other actions across the entire life cycle of plastics. They believe that the best way to address plastic pollution comes from prevention over treating the effects of pollution. The data is two years (2019 and 2020) worth of data collected through global audits that BFFP does trying to understand the sources of plastic waste creation. The information provided by the data set is the country, parent company which created the plastic waste, the number of clean up events and volunteers, and the counts of plastic waste. This was unit-less so I am assuming individual plastic pieces collected at the event. There was a grand total count and multiple columns representing the different types of plastics ranging from the thin material of bags to PVC pipe plastic. Each row seems to represent a country and company combination. There are over 10,000 rows of data.

This data was originally found in multiple folders one each for the years 2019 and 2020. Each folder then had a separate csv file for each country. jonthegeek used map_dfr to combine all the csv files into a single data frame. This function also changed the data types. All were turned into doubles as they were numbers except for the country and parent company as these were characters. He then added a year column as the ideas was to eventually combine the two years together. After cleaning up the names with janitor he did the same thing to the 2020 data. This one had some duplicate columns so he used an if_else function to take care of that. Finally the two years were bound together and some names were changed to make it easier to work with and we now have the data set we will be using.


## API 1 Data Set
This API provided statistics on the countries of the world. By iterating against the countries in our data set we were able to get information on them. The API is unclear on which years the data is from and because of that we are looking at measures that we consider to be more stable at least within the years that our original data is from, 2019 and 2020. Sectors of employment is one of those fields. Other there are other interesting variables in this new data as well including but not limited to endangered species, forested area, and co2 emmissions.

## API 2 Data Set
This dataset was obtained through the Global Footprint Network API by pulling country level data for 2019 and 2020 from the National Footprint and Biocapacity Accounts. The API returns environmental and development indicators for countries and territories, including metrics such as Total Biocapacity and Total Ecological Footprint. After retrieving the data, I combined the yearly responses, reshaped them into a country year format, and matched them to the plastics dataset using country and year. This allowed me to incorporate environmental context into the original dataset and compare plastic collection outcomes with  a nations ecological state.


# Research Questions
## Question 1
How do demographic and economic characteristics of countries along with their geographic location relate to differences in plastic collected during cleanup audits?

## Question 2
Does the economic well being of a country affect the proportion in which they use specific types of plastics?


# Data Summaries and Vizualizations

```{r}
#| label: Libraries
#| message: false

library(tidyverse)
library(gt)
library(httr)
library(stringr)
library(tidyjson)
library(jsonlite)
library(ggridges)
library(RColorBrewer)
library(viridis)
library(ggtext)
library(broom)
library(scales)
library(car)

plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

plastics_clean <- plastics |> 
  mutate(across(empty:pvc, ~replace_na(., 0)),
         total = rowSums(across(c(empty:pvc))),
         .before = num_events)

final_merged <- readRDS("full_table")
```


## Data Summaries
### Summary statistics for each plastic type across all audits
Need to fix the map_dfr superseeded function

### Proportions of Plastics for each Country
this loks like a good one to do the additional styling on 
```{r}
#| label: Data Sumary 2

proportions <- function(df) {
  
  df |>
    select(country:pvc) |>
    map_if(is.numeric, ~.x / df$grand_total) |>
    as.data.frame() 
}

plastics_clean |>
    mutate(year = factor(year)) |>
    group_by(country, year) |>
    summarize(across(empty:total, 
                     ~sum(.x, na.rm = TRUE)),
              .groups = "drop") |>
  proportions() |>
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia",
                        "Nigeria")) |>
  gt() |>
  fmt_percent(columns = empty:pvc) |>
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) |>
  tab_caption("Percentage of each plastic type found in audits from 
              highly populated countries by year") |>
  cols_move_to_end(o) |>
  cols_label(country = "Country", 
             year = "Year",
             empty = "Empty",
             hdpe = "HDPE",
             ldpe = "LDPE",
             o = "Other",
             pet = "PET",
             pp = "PP",
             ps = "PS",
             pvc = "PVC")
```


## Vizualizations

### Ridge Plot Showing Plastic Waste Type Distribution by Contry GDP per Capita
```{r}
#| label: Vizualization 1

final_merged |> 
  filter(year == 2019) |> 
  group_by(country) |> 
  summarize(unknown = sum(empty),
            hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            other = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total = sum(total),
            gdp_per_capita = mean(gdp_per_capita),
            hdi = mean(HDI),
            .groups = "drop") |> 
  mutate(income_group = case_when(
    gdp_per_capita < 1500 ~ "Low Income",
    gdp_per_capita > 12000 ~ "High Income",
    TRUE ~ "Middle Income")) |> 
  pivot_longer(
    cols = unknown:pvc,
    names_to = "plastic_type",
    values_to = "waste") |> 
  filter(waste <= 3500) |> 
  
  ggplot(mapping = aes(x = waste,
                       y = plastic_type,
                       fill = income_group)) +
  geom_density_ridges(alpha = 0.4) +
  scale_fill_viridis(discrete=TRUE) +
  labs(
    title = "Plastic Waste Type Distribution of Countries by GDP per Capita",
    subtitle = "Divided into <span style='color: #6a546f;'>High Income</span>, 
      <span style='color: #b5aa5f;'>Middle Income</span>,
    and <span style='color: #5e8684;'>Low Income</span>",
    x = "Waste Amount",
    y = "",
    alt = "Density ridge plot showing the density of plastic waste audited
    for multiple types of plastic. Colored by GDP per capita category,
    High, Middle, and Low") +
  theme_dark() +
  theme(plot.subtitle =  element_markdown(),
        legend.position = "none") 

```

### Bubble Plot
```{r}
#| label: Vizualization 2
#| warning: false

final_merged |>
  mutate(region = countrycode::countrycode(country,
            origin = "country.name",
            destination = "continent"),
         total_normalized = total / num_events, 
         gdp_trillions = gdp / 1e6) |>
  filter_out(country == "EMPTY",
           total == 0,
           is.na(total),
           is.na(gdp_trillions),
           gdp_trillions == 0, 
           is.na(region)) |>
  group_by(country, region, year) |>
  summarize(total_plastic = sum(total_normalized, na.rm = T), 
            gdp_trillions = mean(gdp_trillions, na.rm = T), 
            Population = mean(Population, na.rm = T)) |>
  ggplot(aes(x = gdp_trillions, y = total_plastic, size = Population, 
             color = region)) +
    geom_point(alpha = 0.5) +
    scale_size(range = c(.1, 15), name ="Population") +
  labs(color = "Region",
       x = "GDP (Trillions)", 
       title = "Plastic Pollution Found in Audits vs. GDP for 2019 and 2020",
       subtitle = "Total Plastic per Cleanup Event",
       y = "",
       alt = "Bubble chart showing plastic items found per cleanup event vs GDP in trillions,
       colored by region and sized by population") +
  coord_cartesian(ylim = c(0, 26000),
                  xlim = c(0,25)) + 
  scale_color_discrete(na.translate = FALSE) +
  scale_x_sqrt() +
  facet_wrap(~ year) +
  theme_light() + scale_color_manual(values = c("Africa" = "mediumspringgreen", "Americas" = "orchid4", 
                                "Asia" = "darkorange", "Europe" = "deeppink", 
                                "Oceania" = "royalblue"))
```

# Data Model

```{r}
#| label: Data Model


model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population = first(na.omit(Population)),
    pop_density = first(na.omit(pop_density)),
    employment_industry = first(na.omit(employment_industry)),
    HDI = first(na.omit(HDI)),
    gdp = first(na.omit(gdp)),
    imports = first(na.omit(imports)),
    infant_mortality = first(na.omit(infant_mortality)),
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_1 <- lm(
  log_total_plastic ~ Population + pop_density + region,
  model_df
)

fit_2 <- lm(
  log_total_plastic ~ Population + pop_density + employment_industry + region,
  model_df
)

fit_3 <- lm(
  log_total_plastic ~ Population + pop_density + HDI + region,
  model_df
)

fit_4 <- lm(
  log_total_plastic ~ Population + pop_density + imports + region,
  model_df
)

fit_5 <- lm(
  log_total_plastic ~ Population + pop_density + infant_mortality + region,
  model_df
)


fit_6 <- lm(
  log_total_plastic ~ Population + pop_density + region + gdp,
  model_df
)

fit_7 <- lm(
  log_total_plastic ~ Population + pop_density + region + gdp + imports,
  model_df
)

adj_rsq <- list(
  summary(fit_1)$adj.r.squared,
  summary(fit_2)$adj.r.squared,
  summary(fit_3)$adj.r.squared,
  summary(fit_4)$adj.r.squared,
  summary(fit_5)$adj.r.squared,
  summary(fit_6)$adj.r.squared,
  summary(fit_7)$adj.r.squared
)

adj_rsq


vif_results <- list(
  vif(fit_1),
  vif(fit_2),
  vif(fit_3),
  vif(fit_4),
  vif(fit_5),
  vif(fit_6),
  vif(fit_7)
)

vif_results

```

```{r}
#| label: Data Model Output
#| warning: false

model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population_millions = first(na.omit(Population)) / 1000000,
    pop_density = first(na.omit(pop_density)),
    gdp_billions = first(na.omit(gdp)) / 1000000000,
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_6 <- lm(
  log_total_plastic ~ Population_millions + pop_density + gdp_billions + region,
  model_df
)

tbl <- tidy(fit_6, conf.int = TRUE) |>
  mutate(
    Predictor = dplyr::recode(term,
      `(Intercept)` = "Intercept",
      `Population_millions` = "Population (millions)",
      `pop_density` = "Population Density",
      `gdp_billions` = "GDP (billions)"
    ),
    Predictor = sub("^region", "Region: ", Predictor)
  ) |>
  select(
    Predictor,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

tbl |>
  gt() |>
  tab_header(
    title = md("**Linear Regression Predicting Logged Plastic Collected by Country, Year**"),
    subtitle = "Response: log(total_plastic)"
  ) |>
  cols_label(
    Predictor = md("**Predictor**"),
    Estimate = md("**Estimate**"),
    `Std. Error` = md("**Std. Error**"),
    `t value` = md("**t value**"),
    `p value` = md("**p value**"),
    `95% CI Lower` = md("**95% CI Lower**"),
    `95% CI Upper` = md("**95% CI Upper**")
  ) |>
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `95% CI Lower`, `95% CI Upper`),
    decimals = 4
  ) |>
  fmt_number(columns = `p value`, decimals = 3) |>
  tab_source_note(
    source_note = md(
       "The response variable was transformed as log(total_plastic) to meet linear regression assumptions"
    )
  )
```

A linear regression model was fit to examine how demographic, economic, and regional factors relate to differences in plastic collected during cleanup audits across countries and years. The model included population (in millions), population density, GDP (in billions), and regions as predictors of the log transformed plastic counts. The results show that population density is significantly associated with plastic collected (p = 0.042), indicating that more densely populated countries tend to have lower levels of plastic collected, holding the other factors constant. Regional differences were the strongest predictors, with countries in Western Asia showing significantly lower levels compared to the reference region (p = 0.001). Total population (p = 0.212) and GDP (p = 0.169) were not statistically significant, which suggests that the economic size does not explain variation in plastic collected. The results show that geographic location and population distribution are more important than economic indicators in explaining differences in plastic collected.